---
title: "Worksheet: EDA Worksheet"
subtitle: "STT 2860 Intro to Data Mgmt & Visualization" 
author: "Dalton Littleton"
date: "`r format(Sys.time(), '%A, %B %d, %Y @ %I:%M %p')`"
output: 
  html_document: 
    theme: yeti
    highlight: textmate
    toc: true
    toc_depth: 4 
---

<hr>

```{r globaloptions, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  comment = NA
)
```

## How We’ll Work Today (Groups + Individual Submission)

* **Use the same groups as Project 1.**
* You are encouraged to **talk, compare code, and help each other debug**, but **each student must submit their own completed worksheet to GitHub** (your own repo).
* If you get stuck, follow this order:

  1. read the error carefully
  2. ask your group
  3. ask me

---

## GitHub Workflow Reminder (Do This All Class)

**Before you start**

* Pull the most recent version (if needed).
* Open the `.Rmd`, knit once to make sure it runs.

**During work**

* Work in small steps.
* After each checkpoint section: **Knit → Commit → Push**

---

## Learning Notes (You will write as you go)

Create short notes while you work. You’ll fill these in at the end.

* **One thing I learned today:**
* **One thing that confused me at first but makes sense now:**
* **One ggplot trick I want to remember:**

---

# Packages Used

```{r packages}
# load all packages used here
library(tidyverse)
library(nycflights13)
library(skimr)
```

---

# Datasets Used

## Popular Kids

The dataset `popular.csv` and you can read about it [HERE](https://stat-jet-asu.github.io/Datasets/InstructorDescriptions/popular.html).

```{r kids-read}
# import the dataset from the web URL
kids <- read_csv("https://aokine.github.io/Data/popular.csv")
```

```{r kids-explore}
# explore the dataset
# Use glimpse()
glimpse(kids)
# Use skim()
skim(kids)
```



## Semmelweis Data

The dataset `semmelweisannualbyclinic.csv` and you can read about Dr. Ignaz Semmelweis [HERE](https://www.pbs.org/newshour/health/ignaz-semmelweis-doctor-prescribed-hand-washing). This dataset contains annual observations for two different clinics that led him to implement his handwashing experiment.

```{r semm-read}
semm <- read_csv("https://aokine.github.io/Data/semmelweisannualbyclinic.csv")
```

```{r semm-explore}
# explore the dataset
glimpse(semm)
skim(semm)
```

---

## Checkpoint 1 (Setup + Datasets)

**Checkpoint task:** Knit your document. If it works, **commit and push** with:

> `worksheet-eda: checkpoint-1 (setup + datasets)`

---

# NFL Players 2016 Dataset

Description: [https://stat-jet-asu.github.io/Datasets/InstructorDescriptions/NFLplayers2016.html](https://stat-jet-asu.github.io/Datasets/InstructorDescriptions/NFLplayers2016.html)

## Load the dataset

```{r nfl-read}
NFL <- read_csv("https://aokine.github.io/Data/NFLplayers2016.csv")
```

## Explore the structure with `skim()`

```{r nfl-skim}
# Use skim()
skim(NFL)
```

---

# Barplots with `geom_bar()`

For `geom_bar()`, we input an `x` variable and the geometry counts how many of each value occurs in that variable, and that becomes the "y" value for the height of the bars.

## Barplot: Position Groups

Use `x = PositionGrp`.

```{r bar-1}
# TODO: Create barplot of PositionGrp
ggplot(NFL, aes(x = PositionGrp)) +
  geom_bar()
```

### Choose-your-own mini task (pick ONE)

1. Change bar color and outline (`fill =`, `color =`) **OR**
2. Flip coordinates (`coord_flip()`) **OR**
3. Make the bars show proportions instead of counts (hint: `aes(y = after_stat(count/sum(count)))`)

```{r bar-1-choice}
# Do ONE of the three options above. Add your code here.
ggplot(NFL, aes(x = PositionGrp)) +
  geom_bar() +
  coord_flip()
```
---

## Fixing category names: Map `PositionGrp` to the y aesthetic

Make `PositionGrp` go on the **y-axis** so labels are easier to read.

```{r bar-2}
# TODO: Make horizontal barplot
ggplot(NFL, aes(y = PositionGrp)) +
  geom_bar()
```
---

## Barplot for Two Variables

Use `PositionGrp` and `Side` as the two variables. Use color/fill to separate groups.

```{r bar-3}
# TODO: Add fill = Side
ggplot(NFL, aes(y = PositionGrp, fill = Side)) +
  geom_bar()
```

### Quick interpretation (1–2 sentences)

In plain language, what does this plot suggest about how offense/defense/special teams are distributed across position groups?
**Answer:**
Offense's are spread put based on position groups while defense has a high count of players in less categories than offense. Special teams was limited to one category and had the least amount of players. 



---

## Checkpoint 2 (Skim + geom_bar)

**Checkpoint task:** Knit your document. If it works, **commit and push** with:

> `worksheet-eda: checkpoint-2 (skim + first barplots)`

---

# Barplots with `geom_col()`

For `geom_col()`, we input an `x` variable and we input a `y` variable that specifies bar heights. That means we usually **summarize first**.

## Count the Position Groups

```{r col-1}
# TODO: Create summarized dataset using count()
position_counts <- NFL %>%
  count(PositionGrp)

position_counts
```

## Plot counts with `geom_col()`

```{r col-2}
# TODO: Plot using geom_col()
ggplot(position_counts, aes(x = PositionGrp, y = n)) +
  geom_col()
```

---

## Rearrange the bars in order using `reorder()`

Change the order of the categories according to their counts (largest to smallest).

```{r col-3}
# TODO: Use reorder() to sort bars
ggplot(position_counts, aes(x = reorder(PositionGrp, -n), y = n)) + 
  geom_col()

```

### Choose-your-own mini task (pick ONE)

1. Add data labels above the bars (hint: `geom_text(aes(label = n), ...)`) **OR**
2. Change to descending order clearly (you might need `-n`) **OR**
3. Show proportions instead of counts (create a `prop = n/sum(n)` column)

```{r col-choice}
# Do ONE of the three options above. Add your code here.
ggplot(position_counts, aes(x = reorder(PositionGrp, -n), y = n)) + 
  geom_col() +
  geom_text(aes(label = n, vjust = -0.5))
```

---

## Checkpoint 3 (Two-variable bars + geom_col)

**Checkpoint task:** Knit your document. If it works, **commit and push** with:

> `worksheet-eda: checkpoint-3 (two-variable bars + geom_col)`

---

# Facets

## Facet Wrap

### Basic Histogram using `Weight`

```{r hist-1}
# TODO: Basic histogram
ggplot(NFL, aes(x = Weight)) +
  geom_histogram()
```

### Histogram with Fill using `PositionGrp` OR `Side`

Choose ONE: `fill = PositionGrp` or `fill = Side`

```{r hist-2}
# TODO: Add fill
ggplot(NFL, aes(x = Weight, fill = PositionGrp)) +
  geom_histogram()
```

### Histogram using `facet_wrap()` with `PositionGrp`

```{r hist-3}
# TODO: facet_wrap by PositionGrp
ggplot(NFL, aes(x = Weight)) +
  geom_histogram() +
  facet_wrap(~PositionGrp)
```

### Histogram using `facet_grid()` with `Side`

```{r hist-4}
# TODO: facet_grid by Side
ggplot(NFL, aes(x = Weight)) +
  geom_histogram() +
  facet_wrap(~Side)
```

### Quick interpretation (2–3 sentences)

Pick one faceted plot and write what pattern you notice.
**Answer:**
The pattern I see with the facet_wrap(Side) is that the it correlates with facet_wrap(PositionGrp) as you can see the spikes of positions and their weights. As well as seeing the pattern in the wight of the players in correlation to their position you can see the same pattern from earlier that the majority to least number of players is offense to defense and then to special teams. 

---

## Checkpoint 4 (Reorder + facets)

**Checkpoint task:** Knit your document. If it works, **commit and push** with:

> `worksheet-eda: checkpoint-4 (reorder + facets)`

---

# Reflection (Individual)

Fill these in before you submit.

1. **One thing I learned today:**
I learned how to use Github with RStudio and how to use commit and push commands. First I have to knit the R Markdown to an HTML then press the Git button in the Git panel and make sure that both .Rmd and .html are visible. Then check the boxes of both the .Rmd and .html and press commit. Enter the “worksheet-eda:...” and press commit again on the pop up. Back at the Git panel in RStudio press “push” and verify on Github that it was done successfully. 

2. **One thing that confused me at first but makes sense now:**
The thing that confused me the most, but makes sense now, was linking my work to Github from RStudio. I now know the steps used with commit and push to send my work to Github successfully. 

3. **One ggplot trick I want to remember:**
One plot trick I want to remember is that “ggplot” is first then the “geom_” code follows and to make sure after each line of code when creating a “ggplot” to add the plus sign. Therefore, the code is connected to each other. Also to remember when to use “%>%” also known as the pipe operator to connect functions. 

4. **Choose-your-own reflection (pick ONE):**

* A. One plot I’m proud of and why
* B. One plot I would redesign and how
* C. One dataset question I’d investigate next

**Answer:**
One plot I'm proud of is ggplot(NFL, aes(x = Weight, fill = PositionGrp)) + geom_histogram() because the histogram shows clear clarification of the position groups as a whole in regards to their weight. The graph itself looks as if the two different face_wrap graphs were combined together. This histogram eliminates the need for the facet_wraps unless you were to need the individual data of either facet_wrap(~PositionGrp) or facet_wrap(~Side).
---

## Final Submission

**Final task:** Knit one last time. Then **commit and push** with:

> `worksheet-eda: final (reflection + cleanup)`

---

---

```{r}
sessionInfo()
```
